---
title: "Assignment 2 MSB1015 Making Multivariate Statistics Reuseable"
author: "Caroline Collins 6192527"
date: "25 September 2019"
output: html_document
---

Project Synopsis

Data analysis is a long process that does not merely involve running a statistical
analysis, but starts with the raw, experimental data that needs to be processed and
cleaned, the actual analysis or modelling, and the creation of visualization of the
results. 
During the analysis you will likely have to play with the many parameters
you have to select, to get meaningful outcomes. All this needs to be documented,
allowing you and others to easily reproduce your results. 
In this assignment you will use Markdown and R to make an interactive notebook where you combine explanation of the steps you perform, computation, and visualization of the results.

Computational Chemistry Background

Harry Wiener realized in 1947 that if physicochemical properties are determined by
the chemical structure, then the latter must be correlated to the first. In that year he
published work where he made a correlation model that linked 

structural features
with boiling points, 

starting a new era in computational chemistry. 
Making such regression models is now routine in the life sciences, and not only for
physicochemical properties but also biochemical properties.
In this assignment you will use SPARQL to query physicochemical properties from
Wikidata for a series of chemical properties and make a (PLS) Partial Least Squares model
(or similar) in an R Markdown notebook with RStudio. 
You will have to use a training-test set split 
(e.g. use 10% or 20% as test set and the remaining fortraining) 
and use cross-validation 
(e.g. Leave-on-Out or Leave-10%-Out) 
to estimate the number of needed latent variables in your predictive model. 
Expected plot scatterplot of your model predictions for the test set of the dependent variable (the modelled property) versus the experimental (real, observed) value.

Useful sources

For general knowledge about pharmacology: Rang & Dale, ISBN
9780702053627, in that edition Chapter 7 en 60 are of interest
For general knowledge about QSAR, e.g.:
Varnek & Tropsha, Chemoinformatics approaches to virtual screening
Ch. 6 of Kahnâs Recent Trends on QSAR in the Pharmaeutical Perceptions
(ebook)
Wiener H. Structural Determination of Paraffin Boiling Points. Journal of the
American Chemical Society. 1947 Jan;69(1):17-20.
R Markdown: https://rmarkdown.rstudio.com/
Cheatsheet: https://www.rstudio.com/wpcontent/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf
PLS package for R: https://cran.r-project.org/web/packages/pls/
Vignette: https://cran.r-project.org/web/packages/pls/vignettes/pls- manual.pdf

The WikiData SPARQL endpoint is https://query.wikidata.org/bigdata/namespace/wdq/sparql 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set up libraries required for the assignment

```{r}
library("rcdk")
library(WikidataQueryServiceR)
```

STEP 1 EXECUTE QUERY
Using the R package WikidataQueryServiceR. Maintaining the SPARQL query syntax.
The first query checks for any entries still waiting for their  boiling point statements to be entered in WikiData.
The second query returns the names and id's of the compounds, their canonical SMILES chemical structure, and their boiling point.
SMILES is a text string that represents the compound structure in 2D. ie. it psecifies elements, connectivity and bond order, but not the 3D position of each atom.
Examples of SMILES:
ethanol CCO
ethylene C=C
acetylene C#C
2-propanol CC(O)C
This third query is set up to deal with the complex structure of statements with units in Wikidata.
'SELECT ?comp ?compLabel ?bp ?bpUnit ?bpUnitLabel WHERE {
  ?comp wdt:P31/wdt:P279* wd:Q41581 ;
        p:P2102 [
          ps:P2102 ?bp ;
          psv:P2102/wikibase:quantityUnit  ?bpUnit
        ] .
  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en". }
}'
```{r}
sparql_query_missingbp <- 'SELECT DISTINCT ?anes ?anesLabel ?CC ?bp
WHERE {
    ?anes wdt:P31/wdt:P279* wd:Q41581 .
    ?anes wdt:P233 ?CC .
    MINUS {?anes wdt:P2102 ?bp .}
    SERVICE wikibase:label { bd:serviceParam wikibase:language "en"}
}'
results_missingbp = query_wikidata(sparql_query_missingbp)
```
```{r}
sparql_query <- 'SELECT DISTINCT ?anes ?anesLabel ?CC ?bp
WHERE {
    ?anes wdt:P31/wdt:P279* wd:Q41581 .
    ?anes wdt:P233 ?CC .
    ?anes wdt:P2102 ?bp .
    SERVICE wikibase:label { bd:serviceParam wikibase:language "en"}
}'
results_bp = query_wikidata(sparql_query)
```
STEP 2 MULTIVARIATE STATISTICS Partial Least Squares vs Multilinear Regression
covariance
normalise
scale
scores and loadings
STEP 2b CROSS-VALIDATION
STEP 3 VISUALISATION Scatterplot
